Thesis Title

### 2022-02-25 chapters/accronims.tex
updates do introduction  - metagenomics

Global Burden of Disease (GBD)
World Health Organization (WHO)
Global Priority Pathogens (GPP)
antimicrobial resistance (AMR)
T2 Magnetic Resonance (T2MR)
polymerase chain reaction (PCR)
pulsed field gel electrophoresis (PFGE)
Multilocus sequence typing (MLST)
basepairs (bp)
Real-time quantitative PCR (qPCR)
Lateral flow immunoassays (LFIA)
ribossomal RNA (rRNA)
PubMed Central\textsuperscript{\small\textregistered} (PMC)
Public health surveillance systems (PHSS)
dideoxynucleotide triphosphate (ddNTP) 
deoxynucleotide triphosphate (dNTP)
single molecule real-time sequencing (SMRT)

>>>>

Global Burden of Disease (GBD)
World Health Organization (WHO)
acute  respiratory  syndrome  coronavirus  2(SARS-CoV-2)
Middle East respiratory syndrome (MERS)
Global Priority Pathogens (GPP)
antimicrobial resistance (AMR)
T2 Magnetic Resonance (T2MR)
polymerase chain reaction (PCR)
pulsed field gel electrophoresis (PFGE)
Multilocus sequence typing (MLST)
basepairs (bp)
Real-time quantitative PCR (qPCR)
Lateral flow immunoassays (LFIA)
ribossomal RNA (rRNA)
PubMed Central\textsuperscript{\small\textregistered} (PMC)
Public health surveillance systems (PHSS)
dideoxynucleotide triphosphate (ddNTP) 
deoxynucleotide triphosphate (dNTP)
single molecule real-time sequencing (SMRT)
Whole genome sequencing (WGS)
ribosomal RNA (rRNA)
Operational Taxonomic Units (OTUs)
Fluorescence \textit{in situ} hybridisation (FISH)

### 2022-02-25 chapters/introduction.tex
updates do introduction  - metagenomics

The Global Burden of Disease (GBD) 2019 study reported that microbial pathogens are responsible for more than 400 million years of life lost annually across the globe, a higher burden than either cancer or cardiovascular disease \citep{vos_global_2020}. In particular, lower respiratory infections, diarrhoeal diseases, HIV/AIDS and tuberculosis were amongst the five leading causes of global total years of life lost. More recently, the COVID-19 pandemic, declared as such by the World Health Organization (WHO) on 11 March 2020 after the emergence and global spread of the severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2), and as of January 2022 has caused more than 5.63 million deaths worldwide \citep{ritchie_coronavirus_2020}, making it one of the deadliest pandemics in history. Coronavirus has been responsible for three of the eighteen major pandemics registered throughout modern history \citep{piret_pandemics_2021}, all occurring after the year 2000. \textit{Yersinia pestis}, responsible for three pandemics of plague, \textit{Vibrio cholerae}, with seven cholera pandemics, and Influenza A virus, the causative agent of five flu pandemics, are responsible for the remaining ones, with Influenza being the only other pathogen with a pandemic registered after 2000. Recent decades have also witnessed the emergence of additional virulent pathogens, including the Ebola virus, West Nile virus, Dengue virus and Zika virus, particularly in lower-income countries.

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and DNA-microarray-based hybridisation technology. These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic.

@@

\subsubsection{Fungal infections} \label{sssec:fungal}

Invasive fungal infections are a prevalent disease in pulmonary and critical care medicine, and they frequently pose diagnostic challenges, with laboratory diagnosis of fungal infections being identified as a key area in need of development of a clinical practise guideline \citep{haydour_diagnosis_2019}. Both serologic and molecular approaches are currently in use, with the first including not only immunodiffusion and complement fixation, both of which remain a cornerstone for fungal diagnostic testing, but also lateral-flow assays for fungal antigen detection. Molecular techniques for fungal identification both from culture and directly from patient specimens include nucleic acid probes, mass spectrometry-based methods, and nucleic acid amplification testing \citep{ramanan_laboratory_2017}. Classic fungal culture continues to be relevant and necessary, with most detection and identification methods requiring growth in culture, from phenotypic identification and staining techniques to nucleic acid hybridisation probes and MALDI-TOF mass spectrometry.

Molecular methods of fungal identification directly from the suspect source are particularly attractive for fungi that fail to produce characteristic morphologic characters in a timely fashion, but molecular databases for fungi tend to lag behind bacterial databases in terms of maturity \citep{ramanan_laboratory_2017}. qPCR testing is one of the most frequently used
molecular methods in clinical microbiology, even days before the onset of positive culture result \citep{kourkoumpetis_polymerase_2012}. The ribosomal RNA (rRNA) gene cluster is frequently used as the target for fungi identification, as well as other targets to evaluate antifungal resistance, guiding therapy, which lowers patient mortality and decreases unnecessary antifungal treatment, improving treatment-associated costs and avoiding toxicity \citep{kourkoumpetis_polymerase_2012}.

As with bacterial infections, typing through MLST is also performed for epidemiological studies, with public schemes being available for seven species: \textit{Aspergillus fumigatus}, \textit{Candida albicans}, \textit{Candida glabrata}, \textit{Candida krusei}, \textit{Candida tropicalis}, \textit{Geotrichum spp.} and \textit{Streptomyces spp} (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018}).

\subsubsection{Parasitic infections} \label{sssec:parasites}

Currently, the detection and diagnosis of parasite infections rely on several laboratory methods in addition to clinical symptoms, clinical history, travel history, and geographic location of the patient. serology-based assays, molecular-based assays and microscopy  \citep{ndao_diagnosis_2009}. The use of microscopy for the morphology-based diagnosis of the disease remains the reference standard in parasitology although identification and matching of the life-cycle stages, despite its low cost, is very challenging even with expert knowledge \citep{blasco-costa_molecular_2016}. LFAIs has been used successfully, overcoming many of the disadvantages of conventional microscopy by providing simple, quick, cost-effective, sensitive and accurate identification of parasites, particularly in low-income countries, but are available for a relatively low number of organisms, low sensitivity and reports of cross-reactivity \citep{momcilovic_rapid_2019, robert-gangneux_epidemiology_2012, rajput_false_2018}.

As observed in the diagnosis of bacterial, viral and fungal infections, molecular methods by nucleic acid amplification through qPCR has been increasingly adopted, particularly through the commercialisation of multiplex with high sensitivity and accuracy \citep{momcilovic_rapid_2019, wong_molecular_2014}. The key limitation of molecular detection is the technological expertise and expense which are usually lacking in the field setting at highly endemic areas.

Currently, public MLST schemes for for epidemiological studies of parasitic infections are available for six organisms: \textit{Blastocystis spp.},\textit{ Clonorchis sinensis}, \textit{Kudoa septempunctata}, \textit{Saprolegnia parasitica}, \textit{Treponema pallidum subsp. pallidum} and \textit{Tricohomas vaginalis} (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018}).

@@

\subsubsection{The Third-generation of DNA sequencing} \label{ssec:3rd_gen_seq}

@@

Whole-genome sequencing (WGS) is becoming one of the most widely used applications of microbial genome sequencing. The major advantage of WGS is to yield all the available DNA information content on isolates in a single rapid step following culture (sequencing without culture will be discussed in the \secref{ssec:metagenomics}). In principle, after obtaining a pure culture, either bacterial (see \secref{sssec:bacterial}), viral (see \secref{sssec:viral}) of fungal (see \secref{sssec:fungal}), the data from sequencing contain all the information currently used for diagnostic and typing needs, and much more, thus opening the prospect for large-scale research into pathogen genotype-phenotype associations from routinely collected data \citep{didelot_transforming_2012}.
The cost of producing massive amounts of information requires a new framework with expert handling and processing of computer-driven genomic information, as well as capable computational infrastructures, but through this technology, researchers and clinicians can obtain the most comprehensive view of genomic information and associated biological implications, transforming clinical diagnostics and the detection and surveillance of outbreaks. \citep{cirulli_uncovering_2010, noauthor_genomic_2019, goodwin_coming_2016}.

@@

@@

Because this approach is PCR-based, it suffers from the same issues described previously for conventional PCR, requiring primer design. Additionally, it must necessarily account for intragenomic variation between 16S gene copies. Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 bp) gene \citep{johnson_evaluation_2019}. The emergence of third generating sequencing technologies (see \secref{ssec:3rd_gen_seq}) allows for this limitation to be overcome but currently, only XXX database includes complete 16S rRNA sequences. %TODO WHAT DB

The internal transcribed spacer (ITS) region of rDNA is the most used barcode to study fungal diversity \citep{schoch_nuclear_2012}.  ITS region contains three partitions: ITS1, 5.8S and ITS2. The length of the ITS sequence is highly variable from one fungal species to another and it is strongly dependent on the primers used to target the DNA sequence. Similarly to 16S rRNA sequencing, second-generation sequencing only allows for the sequencing of one of the ITS regions (ITS1 or ITS2). Although ITS1 outperforms ITS2 in terms of richness, and taxonomic coverage, some taxa are exclusively detected only by ITS2, and vice-versa for ITS1 \citep{mbareche_comparison_2020}. This can also be remedied by third generation sequencing technologies (see \secref{ssec:3rd_gen_seq}). Similarly to 16S, dedicated ITS databases that include near full length sequences for a large number of strains and their taxonomic placements exist, such as unite\footnote{\url{https://unite.ut.ee/}}, GlobalFungi\footnote{\url{https://globalfungi.com/}} and NCBI's ITS project\footnote{\url{https://www.ncbi.nlm.nih.gov/refseq/targetedloci/}} \citep{nilsson_unite_2019, vetrovsky_globalfungi_2020}. Additionally, NCBI also provides the 28S ribosomal RNA project and 18S ribosomal RNA project for fungal metataxonomics.

While viruses are an integral part of the microbiota, no universal viral marker genes are available to perform such taxonomic assignments. As an alternative, broad scope viral targeted sequence capture (TSC) panels offer depletion of background nucleic acids and improve the recovery of viral reads by targeting coding sequence from a multitude viral genera, such as VirCapSeq-VERT Capture Panel\footnote{\url{https://sequencing.roche.com/content/dam/rochesequence/worldwide/resources/brochure-vircapseq-vert-capture-panel-SEQ1000117.pdf}} but do not guarantee the full recovery of the viral genome, and can present biases towards certain genera \citep{schuele_assessment_2020, wylie_enhanced_2015}.

@@

Shotgun metagenomics can offer relatively unbiased pathogen detection and characterisation, potentially able to provide genotyping, antimicrobial resistance and virulence profiling in a single methodological step. This comes with the cost of producing massive amounts of information that require expert handling and processing, as well as capable computational infrastructures. One of the biggest challenges when dealing with metagenomic data is the lack of golden standards \citep{couto_critical_2018, rossen_practical_2018}

@@

\section{Bioinformatic Analysis for Shotgun Metagenomics}

>>>>

The Global Burden of Disease (GBD) 2019 study reported that microbial pathogens are responsible for more than 400 million years of life lost annually across the globe, a higher burden than either cancer or cardiovascular disease \citep{vos_global_2020}. In particular, lower respiratory infections, diarrhoeal diseases, HIV/AIDS and tuberculosis were amongst the five leading causes of global total years of life lost. More recently, the COVID-19 pandemic, declared as such by the World Health Organization (WHO) on 11 March 2020 after the emergence and global spread of the severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2), and as of January 2022 has caused more than 5.63 million deaths worldwide \citep{ritchie_coronavirus_2020}, making it one of the deadliest pandemics in history. Coronavirus has been responsible for three of the eighteen major pandemics registered throughout modern history \citep{piret_pandemics_2021}, all occurring after the year 2000 (see \secref{table:introduction:1}. \textit{Yersinia pestis}, responsible for three pandemics of plague, \textit{Vibrio cholerae}, with seven cholera pandemics, and Influenza A virus, the causative agent of five flu pandemics, are responsible for the remaining ones, with Influenza being the only other pathogen with a pandemic registered after 2000. Recent decades have also witnessed the emergence of additional virulent pathogens, including the Ebola virus, West Nile virus, Dengue virus and Zika virus, particularly in lower-income countries.

% TABLE OR FIGURE HERE
\begin{table}[h!]
\caption{Pandemics and epidemics with at least 1 million death throughout modern history. Adapted from \cite{piret_pandemics_2021}.}\label{table:introduction:1}
\tiny
\begin{tabularx}{\textwidth}{@{}llllrll@{}}
\toprule
\textbf{Years} & \textbf{Pandemics/Epidemic} & \textbf{Pathogens} & \textbf{Source/Vector} & \multicolumn{1}{l}{\textbf{Estimated Death toll (Millions)}} & \textbf{Location} & \textbf{Reference} \\ \midrule
165–180 & Antonine Plague & Smallpox or measles & unkown & 5–10 & Roman Empire &  \\
541–543 & Plague of Justinian\authfn{1} & \textit{Yersinia pestis} & Fleas associated to wild rodents & 15–100 & Europe and West Asia &  \\
735–737 & Japanese smallpox epidemic & Smallpox & uncertain & 2 & Japan &  \\
1347–1351 & Black Death\authfn{1} & \textit{Yersinia pestis} & Fleas associated to wild rodents & 75–200 & Europe, Asia, and North Africa &  \\
1519–1520 & Mexico smallpox epidemic & Smallpox & Spanish conquests & 5-8 & Mexico &  \\
1545–1548 & First Cocoliztli epidemic & \textit{Salmonella enterica}\authfn{3} & Cattle\authfn{3} & 5–15 & Mexico &  \\
1576–1580 & Second Cocoliztli epidemic & \textit{Salmonella enterica}\authfn{3} & Cattle\authfn{3} & 2–2.5 & Mexico &  \\
1629–1631 & Italian plague & \textit{Yersinia pestis} & Fleas associated to wild rodents & 1 & Italy &  \\
1656–1658 & Naples plague & \textit{Yersinia pestis} & Fleas associated to wild rodents & 1.25 & Italy &  \\
1772–1773 & Persian Plague & \textit{Yersinia pestis} & Fleas associated to wild rodents & 2 & Persia &  \\
1817–1824 & First cholera pandemic\authfn{1} & \textit{Vibrio cholerae} & Contaminated water & unkown & Worldwide &  \\
1827–1835 & Second cholera pandemic\authfn{1} & \textit{Vibrio cholerae} & Contaminated water & unkown & Worldwide &  \\
1839–1856 & Third cholera pandemic\authfn{1} & \textit{Vibrio cholerae} & Contaminated water & unkown & Worldwide &  \\
1863–1875 & Fourth cholera pandemic\authfn{1} & \textit{Vibrio cholerae} & Contaminated water & unkown & Worldwide &  \\
1881–1886 & Fifth cholera pandemic\authfn{1} & \textit{Vibrio cholerae} & Contaminated water & unkown & Worldwide &  \\
1885–ongoing\authfn{2} & Third plague\authfn{1} & \textit{Yersinia pestis} & Fleas associated to wild rodents & 12–15 & Worldwide &  \\
1881–1886 & Russian flu\authfn{1} & Influenza A/H3N8\authfn{3} & Avian\authfn{3} & 1 & Worldwide &  \\
1899–1923 & Sixth cholera pandemic\authfn{1} & \textit{Vibrio cholerae} & Contaminated water & unkown & Worldwide &  \\
1918–1919 & Spanish flu\authfn{1} & Influenza A/H1N1 & Avian & 17-100 & Worldwide &  \\
1918–1922 & Russia typhus epidemic & \textit{Rickettsia sp} & lice, fleas & 2-3 & Russia &  \\
1957–1959 & Asian flu\authfn{1} & Influenza A/H2N2 & Avian & 1–4 & Worldwide &  \\
1961-ongoing\authfn{2} & Seventh cholera pandemic\authfn{1} & \textit{Vibrio cholerae} & Contaminated water & unkown & Worldwide &  \\
1968–1970 & Hong Kong flu\authfn{1} & Influenza A/H3N2 & Avian & 1–4 & Worldwide &  \\
1981-ongoing\authfn{2} & HIV/AIDS global epidemic & HIV/AIDS & Chimpanzee & \textgreater{}36 & Wordwide &  \\
2002–2003 & Severe acute respiratory syndrome (SARS)\authfn{1} & SARS-CoV & Bats, palm civets & 0.000774 & Wordwide &  \\
2009–2010 & Swine flu\authfn{1} & \textit{Influenza A/H1N1} & Pigs & 0.15-0.57 & Wordwide &  \\
2015-ongoing\authfn{2} & Middle East respiratory syndrome (MERS)\authfn{1} & MERS-CoV & Bats, dromedary camels & 0.000941 & Wordwide &  \\
2019-ongoing\authfn{2} & COVID-19\authfn{1} & SARS-CoV-2 & Bats, pangolins\authfn{3} & \textgreater{}23 & Wordwide &  \\ \bottomrule
\end{tabularx}

\begin{tablenotes}
\item \authfn{1} Pandemic
\item \authfn{2} Cases still being registered currently
\item \authfn{3} Unsure
\item CoV coronavirus;
\item COVID-19 coronavirus disease 2019
\end{tablenotes}

\end{table}

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and DNA-microarray-based hybridisation technology. These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic. for the purpose of this dissertation work, we'll be focusing on bacterial and viral infections.

@@

@@

\subsubsection{The third-generation of DNA sequencing} \label{ssec:3rd_gen_seq}

@@

Whole-genome sequencing (WGS) is becoming one of the most widely used applications of microbial genome sequencing. The major advantage of WGS is to yield all the available DNA information content on isolates in a single rapid step following culture (sequencing without culture will be discussed in the \secref{ssec:metagenomics}). In principle, after obtaining a pure culture, either bacterial (see \secref{sssec:bacterial}), viral (see \secref{sssec:viral}) of fungal (see \secref{sssec:fungal}), the data from sequencing contain all the information currently used for diagnostic and typing needs, and much more, thus opening the prospect for large-scale research into pathogen genotype-phenotype associations from routinely collected data \citep{didelot_transforming_2012}.
The cost of producing massive amounts of information requires a new framework with expert handling and processing of computer-driven genomic information, as well as capable computational infrastructures, but through this technology, researchers and clinicians can obtain the most comprehensive view of genomic information and associated biological implications, transforming clinical diagnostics and the detection and surveillance of outbreaks. \citep{cirulli_uncovering_2010, nature_reviews_genetics_genomic_2019, goodwin_coming_2016}.

@@

Metataxonomics, also referred to as target metagenomics, takes advantage of marker genes that are ubiquitous to a wide range of taxa and that include: (1) highly conserved regions  to facilitate the design of primers for the amplification of the same genomic regions across many taxa, which is required before sequencing, and (2) hyper-variable
regions that enable differentiation even at a finer taxonomic level, such as species.

@@

Because this approach is PCR-based, it suffers from the same issues described previously for conventional PCR, requiring primer design. Additionally, it must necessarily account for intragenomic variation between 16S gene copies. Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 bp) gene \citep{johnson_evaluation_2019}. The emergence of third generating sequencing technologies (see \secref{ssec:3rd_gen_seq}) allows for this limitation to be overcome but currently, only a few complete 16S rRNA sequences are included in the available databases.

While viruses are an integral part of the microbiota, no universal viral marker genes are available to perform such taxonomic assignments. As an alternative, broad scope viral targeted sequence capture (TSC) panels offer depletion of background nucleic acids and improve the recovery of viral reads by targeting coding sequence from a multitude viral genera, such as VirCapSeq-VERT Capture Panel\footnote{\url{https://sequencing.roche.com/content/dam/rochesequence/worldwide/resources/brochure-vircapseq-vert-capture-panel-SEQ1000117.pdf}} but, besides being limited to the taxa baits present in the panel, do not guarantee the full recovery of the viral genome, and can present biases towards certain genera and can\citep{schuele_assessment_2020, wylie_enhanced_2015}.

Similar approaches can be followed with targeted metagenomics, where specific microbial groups are enriched prior to sequencing. This enrichment is not limited to bait panels, but can be performed through a myriad of techniques. Fluorescence \textit{in situ} hybridisation (FISH) using fluorescently labelled oligonucleotide probes provide a targeted approach by staining specifically the microorganism of interest before enriching them by fluorescence activated cell sorting \citep{grieb_pipeline_2020}. Sequenced PCR products, either of total RNA extraction or through the design of adequate probes, allows the targeted sequencing of RNA-viruses within a sample \citep{desdouits_targeted_2020}.

@@

Shotgun metagenomics, also simply called metagenomics and defined by the sequencing of random  DNA fragments of microbial organisms directly from the biological sample, can offer relatively unbiased pathogen detection and characterisation, potentially able to provide genotyping, antimicrobial resistance and virulence profiling in a single methodological step. It is a promising methodology to obtain very fast results for the identification of pathogens and their virulence and resistance properties directly from samples, without the need for culture. This comes with the cost of producing massive amounts of information that require expert handling and processing, as well as capable computational infrastructures. One of the biggest challenges when dealing with metagenomic data is the lack of golden standards \citep{couto_critical_2018, rossen_practical_2018}.

Compared with metataxonomics, it avoids biases from primer choice and the PCR amplification step, enabling, theoretically, not only the detection of organisms across all domains of life, but as well as their functional characterisation in a clinically relevant timeframe \citep{lopez-labrador_recommendations_2021, }.

However, uneven sequencing depth of different organisms and low depth of coverage per species are drawbacks that limit taxa identification and characterisation, such as the identification of mobile genetic elements carrying potentially important genetic elements. Plasmids carrying resistance genes are an example of this \citep{greninger_challenge_2018}. Furthermore, The non-clonal nature of the organisms within a sample further complicates the differentiation between distinct taxa, particularly if they are closely related organisms. Due to the lack of the selection step of PCR amplification, the amount of DNA from the host if significantly higher, further hindering the recovery of microbial information.

@@

% SCRAPS!!!!
\section{Bioinformatic Analysis for Shotgun Metagenomics}
